### 2015-08-25 manuscript.tex
First commit

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

@@

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Estimating admixture histories is crucial for understanding the genetic diversity we see in present-day populations. Existing allele frequency or phylogeny-based methods are excellent for inferring the existence of admixture or its proportions, but have less power for estimating admixture times. Recently introduced approaches for estimating these times use spatial information from admixed chromosomes, such as the local ancestry or the decay of admixture linkage disequilibrium (ALD).  One popular method, implemented in the programs ALDER and ROLOFF, uses two-locus ALD to infer the time of a single admixture event, but is only able to estimate the time of the most recent admixture event based on this summary statistic.  We derive analytical expressions for the expected ALD in a three-locus system and provide a new statistical method based on these results that is able to resolve more complicated admixture histories. Using simulations, we show how this new statistic behaves on a range of admixture histories. As an example, we also apply our method to the Colombian and Mexican samples from the 1000 Genomes project.

\section*{Introduction}
There are many methods for inferring the presence of admixture, e.g. methods
using simple summary statistics detecting deviations from phylogenetic symmetry
\cite{reich2009reconstructing} [TODO(mason) Also cite patterson paper and paper
from Monty's group], methods estimating admixture proportions with programs such
as Structure \cite{pritchard2000inference}, Admixture \cite{alexander2009fast}
or RFmic [cite] [TODO comment: we might want to make sure to add more citations
here, for example all the ones from our previous papers and the ones from the
ALDER paper. People like to see themselves cited]. However, there has been less
research on estimating admixture times, possibly because such methods require
data which were unavailable until the advent of high-throughput next generation
sequencing. Some recently developed methods use the inferred local ancestry of
sequences to construct admixture tract length distributions [TODO cite Pool Nielsen,
Gravel and yourself]. Over time, recombination is expected to decrease the
average lengths of admixture tracts. The length distribution of admixture tracts
is therefore informative about the time since admixture.  Much of the theory
relating to tracts lengths is based on Fisher's famous theory of junctions
\cite{fisher1949theory} and subsequent work by Stam \cite{stam1980distribution}
and others [TODO add as many citations here as possible]. For example,
\cite{baird2003distribution} first discussed the length distribution of tracts
descended from a single ancestor. These results informed later analyses of
admixture tract length distribution, such as \cite{pool2009inference},
\cite{gravel2012population}, and \cite{liang2014lengths}.
\cite{gravel2012population} also implemented the software program TRACTS, which
estimates admixture histories by fitting the tract length distribution, obtained
by local ancestry inference, to a exponential approximation.

Another approach, which we will follow in this paper, is based on the decay of
admixture linkage disequilibrium (ALD). In a well-mixed, genetically isolated
human populations, linkage disequilibrium decays to zero on a scale of tenths of
centiMorgans. However, when an admixed population is founded, it begins with
large of amount of linkage disequilibrium, which is a result of the allele
frequency differences between the source populations. This occurs even if the LD
in the source populations themselves is negligible. The linkage disequilibrium
in the admixed population then fluctuates in the generations after its founding,
decreasing as a result of drift and recombination, or increasing because of
additional waves of migration. From the LD present in a modern day admixed
population, it is possible to make inferences about the population's admixture
history. This insight was first used in the program ROLLOFF
\cite{moorjani2011history} and was later extended by ALDER
\cite{loh2013inferring}.

@@

ROLLOFF and ALDER use the information contained in pairs of sites by examining
the two-locus linkage disequilibrium between them. Here we extend the theory
underlying the methods in ROLOFF and ADLER to three loci by considering
three-locus LD. There are two ways of measuring the linkage between $n$ loci.
\cite{bennett1952theory} defines $n$-locus linkage in a way that maintains a
geometric decrease of LD each generation as a result of recombination, which is
an important property of two-locus linkage disequilibrium.
\cite{slatkin1972treating} defines $n$-locus LD  to be the $n$-way covariance,
analogously to the property of two locus LD as the covariance in allele
frequency between pairs of loci. For two and three loci, these two definitions
coincide, but for four or more loci, they do not.

In this paper, we will use Bennett and Slatkin's definition of three-locus LD to
examine the the decay of ALD for three sites as a function of the genetic
distance between them. We derive an equation that describes the decay of
three-locus LD under an admixture history with multiple waves of migration. We
then compare the results of coalescent simulations to this equation, and develop
some guidelines for when admixture histories more complex than a single pulse
can be resolved. Finally, we apply our method to the Colombian and Mexican
samples in the 1000 Genomes data set, using the Yoruba samples as a reference.
Fitting a two-pulse model to data, we estimate admixture histories for the two
populations which are qualitatively consistent with the results reported in
\cite{gravel2013reconstructing}.

[TODO Could we say anything here about Neanderthals in Asia?]

% You may title this section "Methods" or "Models".
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis"
\section{Model}
We use a random union of gametes admixture model as described in
\cite{liang2014understanding}, which is an extension of the mechanistic
admixture model formulated by \cite{verdu2011general}. In this model, two or
more source populations contribute migrants to form an admixed population
consisting of $2N$ haploid individuals. Each generation in the admixed
population is formed through the recombination of randomly selected individuals
from the previous generation, with some individuals potentially replaced by
migrants from the source populations. For simplicity, we consider a model with
only two source populations. Furthermore, the first source population only
contributes migrants in the founding generation, $T$. The second source
population contributes migrants in the founding generation and possibly in one
or more generations thereafter. In generation $i$, for $i=T-1,\dots,0$ (before
the present), a fraction $m_i$ of the admixed population is replaced by
individuals from the second source population.

\section{Linkage Disequilibrium and Local Ancestry}
ROLLOFF and ALDER use the standard two-locus measure of LD between a SNP at
positions $x$ and another SNP at position $y$, which is a genetic distance $d$
to the right,

\begin{align}
	D_2(d) = \text{cov}(H_x,H_y),
	\label{D2}
\end{align}

where $H_x$ and $H_y$ represent the haplotype or genotypes of an admixed
chromosome at positions $x$ and $y$. In the case of haplotype data, $H_{i,x}=1$
if the $i^\text{th}$ sample is carrying the derived allele at the SNP at
position $x$, and is otherwise 0. Alternatively, for genotype data, $H_{i,x}$
take on values from $\{0,1/2,1\}$ depending on the number of copies of the
derived allele the $i^\text{th}$ sample is carrying at SNP position $x$.
We consider an additional site at position $z$, which is located a further
genetic distance $d'$ to the right of $y$. The three-loci LD, as defined by as
defined by \cite{bennett1952theory} and \cite{slatkin1972treating}, is given by

\begin{align}
	D_3(d,d')
	= \text{cov}(H_x,H_y,H_z)
	= \mathbb{E}[(H_x-\mathbb{E}H_x)(H_y-\mathbb{E}H_y)(H_z-\mathbb{E}H_z)].
\label{D3}
\end{align}

The LD in an admixed population depends on the genetic differentiation between the source populations and and its admixture history. Let $A_x$ represent the local ancestry at position $x$, with $A_x=1$ if $x$ is inherited from an ancestor in the first source population, and $A_x=0$ if $x$ is inherited from the second source population. We can compute $D_3$ in terms of the three-point covariance function of $A_x$ and so separate out the effects of allele frequencies and local ancestry. Let $H_x = f_x+\delta A_x$, where $f_x$ is the allele frequency of locus $x$ in the first source population and $\delta_x$ is the difference of the allele frequencies of locus $x$ in the two source populations. We now make the assumption that the allele frequencies in the source populations are known and fixed. Equation \ref{D3} then becomes

\begin{align}
	D_3(d,d')
		&= \text{cov}\left(f_x+\delta_xA_x,f_y+\delta_yA_y,f_z+\delta_zA_z\right)\nonumber \\
		&= \delta_x\delta_y\delta_z \text{cov}(A_x,A_y,A_z).
	\label{local_ancestry}
\end{align}
A similar argument shows that $D_2(d)$ is proportional to the two-point covariance function of the local ancestry.

\subsection{Local Ancestry Covariance Functions}
From the above section we see that we can describe the three-point admixture LD in terms of covariances of local ancestry in the three points.  We now expand the covariance in equation \ref{D3} into its component expectations to get
\begin{align*}
	\text{cov}(A_x,A_y,A_z)
		&= \mathbb{E}[A_xA_yA_z]
		-\mathbb{E}[A_xA_y]\mathbb{E}[A_z]
		-\mathbb{E}[A_xA_z]\mathbb{E}[A_y]
		-\mathbb{E}[A_yA_z]\mathbb{E}[A_x]
		+2\mathbb{E}[A_x]\mathbb{E}[A_y]\mathbb{E}[A_z].
\end{align*}
Each one of these expectations on the right-hand side is the probability that one or more sites is inherited from an ancestor from first source population. We organize these products of probabilities in a column vector:
\begin{align*}
	\mathbf{v}_3 &= \left(\begin{array}{l}
	\mathbb{P}\{A_x=A_y=A_z=1\}\\
	\mathbb{P}\{A_y=A_z=0\}\mathbb{P}\{A_x=0\}\\
	\mathbb{P}\{A_x=A_z=0\}\mathbb{P}\{A_y=0\}\\
	\mathbb{P}\{A_x=A_y=0\}\mathbb{P}\{A_z=0\}\\
	\mathbb{P}\{A_x=0\}\mathbb{P}\{A_y=0\}\mathbb{P}\{A_z=0\}
	\end{array}\right),
\end{align*}
so that $\text{cov}(A_x,A_y,A_z) = (1,-1,-1,-1,2)\mathbf{v}_3$. There is one entry in $\mathbf{v}_3$ for each of the five ways in which the three markers at positions $x,y$, and $z$ can arranged on one or more chromosomes. In the founding generation $T$, this column vector is given by $\mathbf{v}_{3(T)} = (1-m_T,(1-m_T)^2,(1-m_T)^2,(1-m_T)^2,(1-m_T)^3)'$. The probabilities for subsequent generations can be found by left-multiplying drift, recombination, and migration matrices:
\begin{align*}
	\mathbf{v}_{3(i)} = \mathbf{D}_i \mathbf{L}\mathbf{U} \mathbf{v}_{3(i-1)},
\end{align*}
The matrices $\mathbf{D}_i$, $\mathbf{L}$, and $\mathbf{U}$ account for the effects of migration, drift, and recombination, respectively. The migration matrix is a diagonal matrix given by
\begin{align*}
	\mathbf{D}_i &= \text{diag}(1-m_i,(1-m_i)^2,(1-m_i)^2,(1-m_i)^2,(1-m_i)^3).
\end{align*}
Its entries are the probabilities that one, two, or three chromosomes in the admixed population will not be replaced by chromosomes from the second source population in generation $i$. The lower triangular drift matrix
\begin{align*}
	\mathbf{L}&=\frac{1}{4N^2}\left( \begin{array}{ccccc}
					 4N^2 		& 0 	& 0 & 0 & 0\\
					 2N 	& 2N(2N-1) & 0 & 0 & 0\\
					 2N 	& 0 & 2N(2N-1) & 0 & 0\\
					 2N 	& 0 & 0 & 2N(2N-1) & 0\\
					 1 & 2N-1 & 2N-1 & 2N-1 & (2N-1)(2N-2)
				\end{array} \right)
\end{align*}
gives the standard Wright-Fisher drift transition probabilities between the states as a function of the population size $2N$. Finally, the upper triangular recombination matrix is determined by the recombination rates between the three sites:
\begin{align*}
	\mathbf{U} &=
		\left( \begin{array}{ccccc}
		e^{-d-d'} & (1-e^{-d})e^{-d'} & (1-e^{-d})(1-e^{-d'}) & e^{-d}(1-e^{-d'}) & 0 \\
		0 & e^{-d'} & 0 & 0 & 1-e^{-d'} \\
		0 & 0 & 1-e^{-d}-e^{-d'}+2e^{-d-d'} & 0 & e^{-d}+e^{-d'}-2e^{-d-d'} \\
		0 & 0 & 0 & e^{-d} & 1-e^{-d} \\
		0 & 0 & 0 & 0 & 1 \end{array} \right)
\end{align*}
The covariance function is then given by
\begin{align}
	\text{cov}(A_x,A_y,A_z) &= \left(1,-1,-1,-1,2\right)\left(\prod_{i=0}^{T-1} \textbf{D}_i \mathbf{L}\mathbf{U}\right)\textbf{v}_{3(0)}.
	\label{cov}
\end{align}
We can obtain an analogous equation for $\text{cov}(A_x,A_y)$, involving the migration, drift, and recombination matrices for two loci:
$$
	\text{cov}(A_x,A_y) = \left(1,-1\right)\left(\prod_{i=0}^{T-1} \textbf{D}_i \mathbf{L}\mathbf{U}\right)\textbf{v}_{2(0)}.
$$

In some cases, equation \ref{cov} simplifies further. In a one-pulse migration model, in which $m_T=M$ and is there after 0, the $\mathbf{D}_i$'s become identity matrices, and we get the closed from expression
$$
	\text{cov}(A_x,A_y,A_z) = M(1-M)(1-2M)\left(1-\frac{1}{2N}\right)^T\left(1-\frac{2}{2N}\right)^Te^{-T(d+d')}.
$$
This is because $(1,-1,-1,-1,2)$ is a left eigenvector of both $\mathbf{L}$ and $\mathbf{U}$, with corresponding eigenvalues $(1-1/2N)(1-2/2N)$ and $\exp(-d-d')$. Note that when $M=0$, the covariance function will be identically 0. Another case is a two pulse model in which we ignore the effects of genetic drift. In this model, admixture only occurs $T$ and $T_2$ generations before the present, so that $m_T=M_1,m_{T'}=M_2$, and all other $m_i$'s are 0. Making the substitution $T_1=T-T_2$, the right hand side of equation \ref{cov} becomes
\begin{align}
		&(1-M_1)(1-M_2)e^{-T_2(d+d')}\left[M_2(1-M_1)^2-2M_2^2(1-M_1)^2+M_1(1-2M_1)e^{-T_1(d+d')}\right.\nonumber\\
		&\ \ \left.-M_1M_2(1-M_1)\left(e^{-M_1d}+e^{-M_1 d'}+\left(1-e^{-d}-e^{-d'}+2e^{-d-d'}\right)^{T_1}\right) \right].
\label {ash_2p}
\end{align}
The corresponding expression for the two-point covariance function is given by
\begin{align}
		(1-M_1)(1-M_2)e^{-T_2d}\left(M_2-M_1M_2+M_1e^{-T_1d}\right),
\label{alder_2p}
\end{align}
which is a mixture of two exponentials.

\section{Weighted Linkage Disequilibrium}
As \cite{loh2013inferring} noted, we cannot use the LD in the admixed population directly, because the allele frequency differences in the source populations can be of either sign. As in As \cite{loh2013inferring} , we solve this problem by computing the product of the values of the three-point linkage disequilibrium coefficient with the product of the allele frequency differences. Using equation \ref{local_ancestry} we obtain
$$
\delta_x\delta_y\delta_z D_3(d,d')=\delta_x^2\delta_y^2\delta_z^2\mathbb{E}[\text{cov}(A_x,A_y,A_z)],
$$
because the local ancestry in the admixed sample is independent of the allele frequencies in the admixed population. For inference purposes, we estimate this function by averaging over triples of SNPs which are separated by distances of approximately $d$ and $d'$. The LD term is estimated from the admixed population, while the $\delta$'s are estimated from reference populations which are closely related to the two source populations. We notice that both this approach, as well as the previous approaches (e.g., \cite{loh2013inferring} ), does not take genetic drift in the source populations after the time of admixture into account, i.e. there is an assumption of both this mehtod and previous methods that the allele frequencies in the ancestral source populations can be approximated well using the allele frequencies in the extant populations.

@@

For a given $d$ and $d'$ the set of SNP triples we use in the estimator for the weighted LD is
$$
	S[d,d'] = \left\{x,y,z: d\leq  x-y < d+1 \text{ and } d' \leq y-z < d'+1 \right\}.
$$
Let $f_x$ be empirical allele frequency in the admixed population. An unbiased estimator of the weighted three-point linkage disequilibrium coefficient is then
\begin{align*}
	\hat{a}[d,d']&=
		\frac{1}{|S[d,d']|}
		\sum_{x,y,z\in S[d,d']}\frac{n\sum_{i=1}^n
			\delta_x\delta_y\delta_z(H_{i,x}-f_x)(H_{i,y}-f_y)(H_{i,z}-f_z)}{(n-1)(n-2)}.
\end{align*}

>>>>

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}

@@

% Please keep the abstract between 250 and 300 words
\section*{Introduction}
Estimating admixture histories is crucial for understanding the genetic diversity we see in present-day populations. Existing allele frequency or phylogeny-based methods are excellent for inferring the existence of admixture or its proportions, but have less power for estimating admixture times. Recently introduced approaches for estimating these times use spatial information from admixed chromosomes, such as the local ancestry or the decay of linkage disequilibrium (LD). We introduce such a method, based on weighted three-locus LD. In comparison to existing methods based on weighted two-locus LD, our new method is able to resolve more complicated admixture histories. Using simulations, we show how this new statistic behaves on a range of admixture histories. We also apply our method to the Colombian and Mexican samples from the 1000 Genomes project and infer admixture histories consistent with those previously reported.
There are many methods for inferring the presence of admixture, e.g. with $f$-statistics \cite{reich2009reconstructing} or by estimating admixture proportions with programs such as Structure \cite{pritchard2000inference} or Admixture \cite{alexander2009fast}. However, there has been less research on estimating admixture times, possibly because such methods require data which was unavailable until the advent of high-throughput next generation sequencing. Some of these methods use the inferred local ancestry of sequences to construct admixture tract length distributions. Over time, recombination is expected to decrease the average lengths of admixture tracts. This was tract length distribution first worked out in the context of junctions \cite{fisher1949theory} and later extended to randomly mating populations by \cite{stam1980distribution}. \cite{baird2003distribution} first discussed the lengths of tracts descended from a single ancestor. These results informed later analyses of admixture tract length distribution, such as \cite{pool2009inference}, \cite{gravel2012population}, and \cite{liang2014lengths}. \cite{gravel2012population} also implemented the software program TRACTS, which estimates admixture histories by fitting the tract length distribution, obtained by local ancestry inference, to a exponential approximation.

Another approach, which we will follow in this paper, is based on the decay of ancestral linkage disequilibrium (LD). In a well-mixed, genetically isolated human populations, linkage disequilibrium decays to zero on a scale of tenths of centiMorgans. However, when an admixed population is founded, it begins with large of amount of linkage disequilibrium, which is a result of the allele frequency differences between the source populations. This occurs even if the LD in the source populations themselves is negligible. The linkage disequilibrium in the admixed population then fluctuates in the generations after its founding, decreasing as a result of drift and recombination, or increasing because of additional waves of migration. From the LD present in a modern day admixed population, it is possible to make inferences about the population's admixture history. This technique was first in the program ROLLOFF \cite{moorjani2011history} and was later extended by ALDER \cite{loh2013inferring}.

@@

ROLLOFF and ALDER use the information contained in pairs of sites by looking at the two-locus linkage disequilibrium between them. We use the information in triples of sites by considering three-locus LD. There are two ways of measuring the linkage between $n$ loci. Two-locus linkage disequilibrium decreases geometrically each generation as a result of recombination. \cite{bennett1952theory} defines $n$-locus linkage in a way that this property is maintained. Another property of two-locus LD is that it is equal to the covariance in the allele frequencies between the two sites. \cite{slatkin1972treating} defines $n$-locus LD analogously. For two and three loci, these two definitions coincide, but for four or more loci, they do not.

In this paper, we will use Bennett and Slatkin's definition of three-locus LD to look at the decay of weighted LD for three sites as a function of the genetic distance between them. We derive an equation that describes the decay of three-locus LD under an admixture history with multiple waves of migration. We then compare the results of coalescent simulations to this equation, and develop some guidelines for when admixture histories more complex than a single pulse can be resolved. Finally, we compute the our method for the Colombian and Mexican samples in the 1000 Genomes data set, using the Yoruba samples as a reference. Fitting a two-pulse model to data, we estimate admixture histories for the two populations which are qualitatively consistent with the results reported in \cite{gravel2013reconstructing}.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section{Model}
We use the same random union of gametes admixture model as in \cite{liang2014understanding}, which is itself an extension of the mechanistic admixture model formulated by \cite{verdu2011general}. In this model, two (or more) source populations contribute migrants to form an admixed population consisting of $2N$ haploid individuals. Each generation in the admixed population is formed through the recombination of randomly selected individuals from the previous generation, with some individuals potentially replaced by migrants from the source populations. For simplicity, we consider a model with only two source populations. Furthermore, the first source population only contributes migrants in the founding generation, $T$. The second source population contributes migrants in the founding generation and possibly in one or more generations thereafter. In generation $i$, for $i=T-1,\dots,0$ (before the present), a fraction $m_i$ of the admixed population is replaced by individuals from the second source population.

\section{Linkage Disequilibrium and Local Ancestry}
ROLLOFF and ALDER use the standard two-locus measure of LD between a SNP at positions $x$ and another SNP at position $y$, which is a genetic distance $d$ to the right, 
\begin{align}
D_2(d) = \text{cov}(H_x,H_y),
\label{D2}
\end{align}
where $H_x$ and $H_y$ represent the haplotype or genotypes of an admixed chromosome at positions $x$ and $y$. In the case of haplotype data, $H_{i,x}=1$ if the $i^\text{th}$ sample is carrying the derived allele at the SNP at position $x$, or is 0 otherwise. Alternatively, for genotype data, $H_{i,x}$ take on values from $\{0,1/2,1\}$ depending on the number of copies of the derived allele the $i^\text{th}$ sample is carrying at the SNP position $x$. 
We consider an additional site at position $z$, which is located a further genetic distance $d'$ to the right of $y$. The three-loci LD, as defined by as defined by \cite{bennett1952theory} and \cite{slatkin1972treating}, is given by
\begin{align}
	D_3(d,d')
	= \text{cov}(H_x,H_y,H_z)
	= \mathbb{E}[(H_x-\mathbb{E}H_x)(H_y-\mathbb{E}H_y)(H_z-\mathbb{E}H_z)].
\label{D3}
\end{align}
The LD in an admixed population depends on the genetic differentiation between the source populations and and its admixture history. Let $A_x$ represent the local ancestry at position $x$, with $A_x=1$ if $x$ is inherited from an ancestor in the first source population, and $A_x=0$ if $x$ is inherited from the second source population. We can compute the expectation of $D_3$ in terms of the three-point covariance function of $A_x$ and so separate out the effects of allele frequencies and local ancestry. We make the assumption that the alleles in the source populations are independent, so that  
\begin{align*}
	\text{cov}\left(H_x,H_y,H_z\right)
		&= \text{cov}\left(\mathbb{E}[H_x|A_x],\mathbb{E}[H_y|A_y],\mathbb{E}[H_z|A_z]\right).
\end{align*}
The background LD in unadmixed human populations decays to zero on a scale of tenths of centiMorgans, so this approximation is appropriate when $d$ and $d'$ are both larger than 0.5 cM. The conditional expectations above are the allele frequencies at each site in the admixed population, conditional on the local ancestry. These are given by $\mathbb{E}[H_x|A_x] = F_x+\delta A_x$, where $F_x$ is the allele frequency of locus $x$ in the first source population and $\delta_x$ is the difference of the allele frequencies of locus $x$ in the two source populations. Equation \ref{D3} becomes 
\begin{align}
	D_3(d,d')
		&= \text{cov}\left(f_x+\delta_xA_x,f_y+\delta_yA_y,f_z+\delta_zA_z\right)\nonumber \\	
		&= \delta_x\delta_y\delta_z \text{cov}(A_x,A_y,A_z).
	\label{local_ancestry}
\end{align}
A similar argument shows that $D_2(d)$ is proportional to the two-point covariance function of the local ancestry.

\subsection{Local Ancestry Covariance Functions}
If we take genetic drift into account, the three-point covariance function is random. To compute its expectation, we multiply out the covariance in equation \ref{D3} to get
\begin{align*}
	\mathbb{E}[\text{cov}(A_x,A_y,A_z)]
		&= \mathbb{E}[A_xA_yA_z]
		-\mathbb{E}[A_xA_y]\mathbb{E}[A_z]
		-\mathbb{E}[A_xA_z]\mathbb{E}[A_y]
		-\mathbb{E}[A_yA_z]\mathbb{E}[A_x]
		+2\mathbb{E}[A_x]\mathbb{E}[A_y]\mathbb{E}[A_z].
\end{align*}
Each one of these expectations on the right-hand side is the probability that one or more sites is inherited from an ancestor from first source population. We organize these products of probabilities in a column vector:
\begin{align*}
	\mathbf{v}_3 &= \left(\begin{array}{l}
	\mathbb{P}\{A_x=A_y=A_z=1\}\\ 
	\mathbb{P}\{A_y=A_z=0\}\mathbb{P}\{A_x=0\}\\
	\mathbb{P}\{A_x=A_z=0\}\mathbb{P}\{A_y=0\}\\
	\mathbb{P}\{A_x=A_y=0\}\mathbb{P}\{A_z=0\}\\ 
	\mathbb{P}\{A_x=0\}\mathbb{P}\{A_y=0\}\mathbb{P}\{A_z=0\}
	\end{array}\right),
\end{align*}
so that $\text{cov}(A_x,A_y,A_z) = (1,-1,-1,-1,2)\mathbf{v}_3$. There is one entry in $\mathbf{v}_3$ for each of the five ways in which the three markers at positions $x,y$, and $z$ can arranged on one or more chromosomes. In the founding generation $T$, this column vector is given by $\mathbf{v}_{3(T)} = (1-m_T,(1-m_T)^2,(1-m_T)^2,(1-m_T)^2,(1-m_T)^3)'$. The probabilities for subsequent generations can be found by left-multiplying drift, recombination, and migration matrices:
\begin{align*}
	\mathbf{v}_{3(i)} = \mathbf{D}_i \mathbf{L}\mathbf{U} \mathbf{v}_{3(i-1)},
\end{align*}
The matrices $\mathbf{D}_i$, $\mathbf{L}$, and $\mathbf{U}$ account for the effects of migration, drift, and recombination, respectively. The migration matrix is a diagonal matrix given by 
\begin{align*}
	\mathbf{D}_i &= \text{diag}(1-m_i,(1-m_i)^2,(1-m_i)^2,(1-m_i)^2,(1-m_i)^3).
\end{align*}
Its entries are the probabilities that one, two, or three chromosomes in the admixed population will not be replaced by chromosomes from the second source population in generation $i$. The lower triangular drift matrix 
\begin{align*}
	\mathbf{L}&=\frac{1}{4N^2}\left( \begin{array}{ccccc}
					 4N^2 		& 0 	& 0 & 0 & 0\\
					 2N 	& 2N-1 & 0 & 0 & 0\\
					 2N 	& 0 & 2N-1 & 0 & 0\\
					 2N 	& 0 & 0 & 2N-1 & 0\\
					 1 & 2N-1 & 2N-1 & 2N-1 & (2N-1)(2N-2)
				\end{array} \right)
\end{align*}
gives the standard Wright-Fisher drift transition probabilities between the states as a function of the population size $2N$. Finally, the upper triangular recombination matrix is determined by the recombination rates between the three sites:
\begin{align*}
	\mathbf{U} &=
		\left( \begin{array}{ccccc}
		e^{-d-d'} & (1-e^{-d})e^{d'} & (1-e^{-d})(1-e^{-d'}) & e^{-d}(1-e^{-d'}) & 0 \\
		0 & e^{-d'} & 0 & 0 & 1-e^{-d'} \\
		0 & 0 & 1-e^{-d}-e^{-d'}+2e^{-d-d'} & 0 & e^{-d}+e^{-d'}-2e^{-d-d'} \\ 
		0 & 0 & 0 & e^{-d} & 1-e^{-d} \\ 
		0 & 0 & 0 & 0 & 1 \end{array} \right)
\end{align*}
The covariance function is then given by
\begin{align}
	\mathbb{E}[\text{cov}(A_x,A_y,A_z)] &= \left(1,-1,-1,-1,2\right)\left(\prod_{i=0}^{T-1} \textbf{D}_i \mathbf{L}\mathbf{U}\right)\textbf{v}_{3(0)}.
	\label{cov}
\end{align} 
We can obtain an analogous equation for $\text{cov}(A_x,A_y)$, involving the migration, drift, and recombination matrices for two loci:
$$
	\mathbb{E}[\text{cov}(A_x,A_y)] = \left(1,-1\right)\left(\prod_{i=0}^{T-1} \textbf{D}_i \mathbf{L}\mathbf{U}\right)\textbf{v}_{2(0)}.
$$

In some cases, equation \ref{cov} simplifies further. In a one-pulse migration model, in which $m_T=M$ and is there after 0, the $\mathbf{D}_i$'s become identity matrices, and we get the closed from expression 
$$
	\mathbb{E}[\text{cov}(A_x,A_y,A_z)] = M(1-M)(1-2M)\left(1-\frac{1}{2N}\right)^T\left(1-\frac{2}{2N}\right)^Te^{-T(d+d')}.
$$
This is because $(1,-1,-1,-1,2)$ is a left eigenvector of both $\mathbf{L}$ and $\mathbf{U}$, with corresponding eigenvectors $(1-1/2N)(1-2/2N)$ and $\exp(-d-d')$. Note that when $M=0$, the covariance function will be identically 0. Another case is a two pulse model in which we ignore the effects of genetic drift. In this model, admixture only occurs $T$ and $T_2$ generations before the present, so that $m_T=M_1,m_{T'}=M_2$, and all other $m_i$'s are 0. Making the substitution $T_1=T-T_2$, the right hand side of equation \ref{cov} becomes 
\begin{align}
		&(1-M_1)(1-M_2)e^{-T_2(d+d')}\left[M_2(1-M_1)^2-2M_2^2(1-M_1)^2+M_1(1-2M_1)e^{-T_1(d+d')}\right.\nonumber\\
		&\ \ \left.-M_1M_2(1-M_1)\left(e^{-M_1d}+e^{-M_1 d'}+\left(1-e^{-d}-e^{-d'}+2e^{-d-d'}\right)^{T_1}\right) \right].
\label {ash_2p}
\end{align} 
The corresponding expression for the two-point covariance function is given by 
\begin{align}
		(1-M_1)(1-M_2)e^{-T_2d}\left(M_2-M_1M_2+M_1e^{-T_1d}\right),
\label{alder_2p}
\end{align} 
which is a mixture of two exponentials. The relative complexity of equation \ref{ash_2p} is actually a feature, as it makes detecting the presence of the second pulse of admixture easier.
\section{Weighted Linkage Disequilibrium}
As \cite{loh2013inferring} noted, we cannot use the LD in the admixed population directly, because the allele frequency differences in the source populations can be of either sign. However, if we compute expectation of the product of the LD with the product of the allele frequency differences, using equation \ref{local_ancestry} we obtain
$$
\mathbb{E}\left[\delta_x\delta_y\delta_z D_3(d,d')\right]=\mathbb{E}[\delta_x^2\delta_y^2\delta_z^2]\mathbb{E}[\text{cov}(A_x,A_y,A_z)],
$$
because the local ancestry in the admixed sample is independent of the allele frequencies in the admixed population. This expectation of the weighted LD is non-zero, and can be estimated by aggregating over triples of SNPs which are separated by distances of approximately $d$ and $d'$. The LD term can be estimated from the admixed population, while the $\delta$'s can be estimated from reference populations which are closely related to the two source populations.

@@

For a given $d$ and $d'$ the set of SNP triples we use in the estimator for the weighted LD is 
$$
	S[d,d'] = \left\{x,y,z: d\leq  x-y < d+1 \text{ and } d' \leq y-z < d'+1 \right\}.
$$
Let $w_x$ be the difference in the empirical allele frequencies in two reference populations and let $f_x$ be empirical allele frequency in the admixed population. An unbiased estimator of the weighted LD is
\begin{align*}
	\hat{a}[d,d']&=
		\frac{1}{|S[d,d']|}
		\sum_{x,y,z\in S[d,d']}\frac{n\sum_{i=1}^n
			w_a w_b w_c(H_{i,x}-f_x)(H_{i,y}-f_y)(H_{i,z}-f_z)}{(n-1)(n-2)}.
\end{align*}

\section{Algorithm}
Directly computing $\hat{a}[d,d']$ over the set $d,d'\in \{0,1,\dots,P\}^2$ would be cubic in the number of segregating sites, but as is the case with ALDER, we can use using a fast Fourier transform (FFT) to approximate $\hat{a}$, giving an algorithm whose run-time is instead linear in the number of segregating sites. We first rearrange $\hat{a}$ to get
\begin{align*}
	\hat{a}[d,d']
		&=\frac{n}{(n-1)(n-2)}
		\frac{\sum_{i=1}^n\sum_{x,y,z\in S[d,d']}
			\delta_x\delta_y\delta_z(H_{i,x}-f_x)(H_{i,y}-f_y)(H_{i,z}-f_z)}
		{\sum_{x,y,z\in S[d,d']}1},
\end{align*}
and define sequences $b_i[d]$ and $c[d]$ by binning the data and then doubling the length by padding with $P$ zeros,
\begin{align*}
	b_i[d] &= \left\{
     \begin{array}{ll}
       \sum_{x: d\leq \lfloor{x}\rfloor < d+1}\delta_x(H_{i,x}-f_x) & : 0\leq d < P\\
       0 & : P\leq d <2P
     \end{array}
   \right.\\
  	c[d] &= \left\{
     \begin{array}{ll}
       |\{x: d\leq \lfloor{x}\rfloor < d+1\}| & : 0\leq d < P\\
       0 & : P\leq d <2P
     \end{array}
   \right.
\end{align*}
We can approximate $|S[d,d']|$ and the $n$ sums in the numerator of $\hat{a}[d,d']$ in terms of convolutions of these sequences:
$$
	|S[d,d']| \approx \sum_{w=0}^P c[w]c[w+d]c[w+d+d']
$$
$$
	\sum_{x,y,z\in S[d,d']}\delta_x\delta_y\delta_z(H_{i,x}-f_x)(H_{i,y}-f_y)(H_{i,z}-f_z)
		\approx
	\sum_{w=0}^P b_i[w]b_i[w+d]b_i[w+d+d'].
$$
These convolutions can be efficiently computed with an FFT, since under a two-dimensional discrete Fourier transform, 
$$
	\sum_{w=0}^P b_i[w]b_i[w+d]b_i[w+d+d'] \leftrightarrow B_i[j]\bar{B_i}[k]B_i[k-j],
$$
where $B_i$ is the (one-dimensional) discrete Fourier transform of $b$, and $B_i[-j]$ is the $j^\text{th}$ to last most element of $B_i$. Summing over $i$ and taking the inverse discrete Fourier transform, we can approximate the discrete Fourier transform of numerator of  $\hat{a}$. We use the same method applied to $c$ to approximate the denominator of $\hat{a}$. Because the number of bins is generally much less than the number of segregating sites, the rate-limiting step of this algorithm is the binning step to form $c$ and the $b_i$'s, which is $O(S_n)$, rather than the FFTs, which are $O(P^2\log(P))$.
